\texttt{nuance}: Detection of planetary transits in the presence of correlated noise

### 2023-01-23 simple/latex/ms.tex
simulated principle dataset description

% Define document class
\documentclass{aastex631}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{bm}
\usepackage{xcolor}

@@

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.75\linewidth]{../figures/principle_linear_search.pdf}
        \caption{Principle and output of the \textit{linear search}. First, a set of durations and depths $\set{T_i, D_j}_{i,j}$ is generated. For each pair of index $(i,j)$, the log likelihood $\ln p(\bm{f} \vert T_i ,D_j, \Delta_{i,j})$ is computed, using the parameters from \autoref{eq:ls} and the expression of \autoref{eq:linear_search_ll}. This process yields the grid of log-likelihoods $\ln\mathcal{L}$ (bottom plot), as well as the $\set{\Delta_{i,j}, \sigma_{i,j}}_{i, j}$ transit depths and errors.}
        \label{fig:linear_search}
    \end{centering}
\end{figure}

@@

\begin{figure}[H]
    \begin{centering}
        \gridline{\fig{../figures/principle_periodic_0.pdf}{0.72\linewidth}{\vspace{-0.5cm}(a) $\ln \mathcal{L}$}}
        \vspace{-0.6cm}
        \gridline{\fig{../figures/principle_periodic_1.pdf}{0.6\linewidth}{\vspace{-0.5cm}(b) $\mathcal{P}(P)$}}
        \caption{On the example dataset, this figure shows how the \textit{periodic search} works at different periods $P$ (including the true period $P = 0.7$ days). Given $P$ and $T_0=0$, the log-likelihood $\ln\mathcal{L}$ shown in (a) is phase-folded and interpolated onto a common grid of phases shown in (b). As an example, the white lines in (a) mark the edges of each fold for a period of $P=0.5$ days, and the white crosses the epochs $\set{T_k}_{k\in\mathbb{K}}$ matching a particular phase in the grid (reported in (a) and (b)) on which the corresponding $\set{\ln \mathcal{L}_k}_{k\in\mathbb{K}}$ are interpolated and combined using \autoref{eq:comb}. To allow the use of efficient matrix computations, this is done for every duration $\set{D_i}_i$ so that $\mathcal{P}$ is computed on the full grid $\set{D_i, \phi_i}_{i, j}$ at once. We understand from plots in (a) that a different choice of epoch $T_0$ may only shift the results in phase but do not affect the values of $\mathcal{P}$. For this reason, computing $\mathcal{P}$ for $T_0=0$ is sufficient. We also notice how the maximum value of $\mathcal{P}(P=0.35)$ is lower than for $P=0.7$ days, a result of combining the log-likelihoods using \autoref{eq:comb} instead of \autoref{eq:attempt}, in favor of individual transits matching a common depth $\Delta$.}
        \label{fig:periodic_search}
    \end{centering}
\end{figure}

@@

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.65\linewidth]{../figures/principle_Q.pdf}
        \caption{}
        \label{fig:periodogram}
    \end{centering}
\end{figure}

@@

This paper relies on simulated light curves featuring transit signals, instrumental signals and correlated noises such as stellar variability. Let such a flux $f$ of a star simulated at time $t$ be sampled and aranged in the $(1\times N)$ column-vector $\bm{f}$ associated to the column-vector of times $\bm{t}$. An example of such signal is simulated and shown in \autoref{fig:principle_dataset}.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.8\linewidth]{../figures/principle_dataset_decomposed.pdf}
        \caption{}
        \label{fig:principle_dataset}
    \end{centering}
\end{figure}
In what follows, we describe the models used to simulate the components of this signal, i.e. the periodic transit $\mathcal{T}$, the instrumental systematics $\mathcal{S}$ and the correlated noise $\mathcal{V}$, such that
$$
    f(t) = \textcolor{gray}{\mathcal{T}}(t) %+ \textcolor{Gray}{\mathcal{S}}(t) + \textcolor{Gray}{\mathcal{V}}(t)
$$
computed for $N=2880$ times $t$ corresponding to an observation of 4 days with an exposure time of 2 minutes. 
\subsection{Transit signal}
We simulate the periodic transit signal $\mathcal{T}$ using the simple model described in \cite{protopapas} where a transit of period $P$, epoch $T_0$, duration $D$ and unitary depth observed at time $t$ is given by
\begin{equation}
    \begin{gathered}
        \mathcal{T}_c(t, P, T_0, D) = \frac{1}{2}\tanh\left(c\left(\theta - \frac{1}{2}\right)\right) - \frac{1}{2}\tanh\left(c\left(\theta + \frac{1}{2}\right)\right) \\
        \text{with}\quad\theta = \frac{P}{\pi  D}\sin\left(\frac{\pi(t-T_0)}{P}\right)
    \end{gathered}
\end{equation}
where the dimensionless parameter c controls the roundness of the transit depth ($c\gg1$ corresponding to a box-shaped transit as shown in \autoref{fig:protopapas}). This analytical model is fully empirical but easily differentiable. From this expression, we get the signal of the single non-periodic transit of unitary depth 
\begin{equation}\label{eq:protopapas_single}
    \tau_c(t, T_0, D) = \lim_{P\to \infty} \mathcal{T}(t, P, T_0, D) = -\frac{1}{2}\tan\left(\frac{c}{2}(1-T_0+t)\right) -\frac{1}{2}\tan\left(\frac{c}{2}(1+T_0-t)\right)
\end{equation}
with epoch $T_0$ and duration $D$.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.9\linewidth]{../figures/protopapas.pdf}
        \caption{Simulations of the single non-periodic transit signal $\tau_c(t, T_0, D)$ (\autoref{eq:protopapas_single}) for different values of $c$}
        \label{fig:protopapas}
    \end{centering}
\end{figure}
In our study, we simulate all transit signals with $c=12$. The periodic transit signal $\mathcal{T}$ seen in \autoref{fig:simu} corresponds to $\mathcal{T} = \mathcal{T}_{c=12}(t, P=0.7, T_0=0.2, D=0.05)$, all parameters in unit of \textit{days}.

\subsection{Instrumental signals}
We simulate instrumental signals as a linear model of $M$ explanatory variables arranged in the $(M\times N)$ design matrix $\bm{X}$. Hence,
$$
    \mathcal{S} \sim \bm{\mathcal{S}} = \bm{w}\bm{X}
$$
where the vector $\bm{w}$ are linear coefficients of the model. The simulated flux shown in \autoref{fig:simu} contains a linear model where the $M=4$ columns of the design matrix $\bm{X}$ are given by $\bm{X}_{i} = \bm{t}^i$ (i.e. $\bm{X}$ is the Vandermonde matrix order $3$ of time $t$) and $\bm{w} = [1.0\quad0.0005\quad\min0.0002\quad-0.0005].$

\subsection{Correlated noise}
We generate stellar variability thanks to the \texttt{tinygp}\footnote{\href{https://github.com/dfm/tinygp}{https://github.com/dfm/tinygp}} Python package, providing an implententation of the scalable GP method from \citealt{celerite2} powered by 
\texttt{JAX}\footnote{\href{https://github.com/google/jax}{https://github.com/google/jax}}.
\section{Combining transits}\label{combining_transits}

>>>>

% Define document class
\documentclass{aastex631}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{bm}
\usepackage[dvipsnames]{xcolor}

@@

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.9\linewidth]{../figures/principle_linear_search.pdf}
        \caption{Principle and output of the \textit{linear search}. First, a set of durations and depths $\set{T_i, D_j}_{i,j}$ is generated. For each pair of index $(i,j)$, the log likelihood $\ln p(\bm{f} \vert T_i ,D_j, \Delta_{i,j})$ is computed, using the parameters from \autoref{eq:ls} and the expression of \autoref{eq:linear_search_ll}. This process yields the grid of log-likelihoods $\ln\mathcal{L}$ (bottom plot), as well as the $\set{\Delta_{i,j}, \sigma_{i,j}}_{i, j}$ transit depths and errors.}
        \label{fig:linear_search}
    \end{centering}
\end{figure}

@@

\begin{figure}[H]
    \begin{centering}
        \gridline{\fig{../figures/principle_periodic_0.pdf}{0.9\linewidth}{\vspace{-0.5cm}(a) $\ln \mathcal{L}$}}
        \vspace{-0.6cm}
        \gridline{\fig{../figures/principle_periodic_1.pdf}{0.7\linewidth}{\vspace{-0.5cm}(b) $\mathcal{P}(P)$}}
        \caption{On the example dataset, this figure shows how the \textit{periodic search} works at different periods $P$ (including the true period $P = 0.7$ days). Given $P$ and $T_0=0$, the log-likelihood $\ln\mathcal{L}$ shown in (a) is phase-folded and interpolated onto a common grid of phases shown in (b). As an example, the white lines in (a) mark the edges of each fold for a period of $P=0.5$ days, and the white crosses the epochs $\set{T_k}_{k\in\mathbb{K}}$ matching a particular phase in the grid (reported in (a) and (b)) on which the corresponding $\set{\ln \mathcal{L}_k}_{k\in\mathbb{K}}$ are interpolated and combined using \autoref{eq:comb}. To allow the use of efficient matrix computations, this is done for every duration $\set{D_i}_i$ so that $\mathcal{P}$ is computed on the full grid $\set{D_i, \phi_i}_{i, j}$ at once. We understand from plots in (a) that a different choice of epoch $T_0$ may only shift the results in phase but do not affect the values of $\mathcal{P}$. For this reason, computing $\mathcal{P}$ for $T_0=0$ is sufficient. We also notice how the maximum value of $\mathcal{P}(P=0.35)$ is lower than for $P=0.7$ days, a result of combining the log-likelihoods using \autoref{eq:comb} instead of \autoref{eq:attempt}, in favor of individual transits matching a common depth $\Delta$.}
        \label{fig:periodic_search}
    \end{centering}
\end{figure}

@@

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.8\linewidth]{../figures/principle_Q.pdf}
        \caption{}
        \label{fig:periodogram}
    \end{centering}
\end{figure}

@@

\definecolor{blue_S}{rgb}{0.12156862745098039, 0.4666666666666667, 0.7058823529411765}
\definecolor{red_V}{rgb}{0.8392156862745098, 0.15294117647058825, 0.1568627450980392}

% This paper relies on simulated light curves featuring periodic transit signals, instrumental signals and correlated noises such as stellar variability. To do so, we draw $\bm{f}$ from a Gaussian Process of mean $\bm{\mu}$ with $\mu_i = {\color{Gray}{\mathcal{T}}}(t_i) + {\color{blue_S}{\mathcal{S}}}(t_i)$, where and covariance matrix $\bm{C}$ with $C_{i, j} = {\color{red_V}\nu}(t_i, t_j)$

Let $f$ be the simulated flux of a star sampled and aranged in the $(1\times N)$ column-vector $\bm{f}$ associated to the column-vector of times $\bm{t}$. We construct $\bm{f}$ such that
$$
    \bm{f} \sim \mathcal{N}(\bm{\mu}, \bm{C})
$$
The mean $\bm{\mu}$ is built such that $\mu_i = \mathcal{T}(t_i) + \mathcal{S}(t_i)$ where $\mathcal{T}$ is a periodic transit signal and $\mathcal{S}$ an instrumental signal. The covariance matrix $\bm{C}$ is built such that $C_{i, j} = \nu(t_i, t_j)$ where $\nu$ is a covariance function accounting for correlated noise in the form of stellar variabilty with added white noise. An example of such signal is simulated and shown in \autoref{fig:app_principle_dataset}.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.8\linewidth]{../figures/principle_dataset_decomposed.pdf}
        \caption{Example dataset sampled at $N=2880$ times corresponding to an observation of 4 days with an exposure time of 2 minutes. The mean of this signal consists in a periodic transit signal plus instrumental signals. Stellar variability and white noise are simulated by modeling the covariance matrix of the signal with Gaussian Processes .}
        \label{fig:app_principle_dataset}
    \end{centering}
\end{figure}
In what follows, we describe the models used to simulate the components of the signal $f$, i.e. the periodic transit $\mathcal{T}$, the instrumental systematics $\mathcal{S}$ and the covariance function $\nu$ describing stellar variability and white noise.

\subsection{Transit signal $\mathcal{T}$}
We simulate the periodic transit signal $\mathcal{T}$ using the simple model described in \cite{protopapas} where a transit of period $P$, epoch $T_0$, duration $D$ and unitary depth observed at time $t$ is given by
\begin{equation}
    \begin{gathered}
        \mathcal{T}_c(t, P, T_0, D) = \frac{1}{2}\tanh\left(c\left(\theta - \frac{1}{2}\right)\right) - \frac{1}{2}\tanh\left(c\left(\theta + \frac{1}{2}\right)\right) \\
        \text{with}\quad\theta = \frac{P}{\pi  D}\sin\left(\frac{\pi(t-T_0)}{P}\right)
    \end{gathered}
\end{equation}
where the dimensionless parameter c controls the roundness of the transit depth ($c\gg1$ corresponding to a box-shaped transit as shown in \autoref{fig:protopapas}). This analytical model is fully empirical but easily differentiable. From this expression, we get the signal of the single non-periodic transit of unitary depth 
\begin{equation}\label{eq:protopapas_single}
    \tau_c(t, T_0, D) = \lim_{P\to \infty} \mathcal{T}(t, P, T_0, D) = -\frac{1}{2}\tan\left(\frac{c}{2}(1-T_0+t)\right) -\frac{1}{2}\tan\left(\frac{c}{2}(1+T_0-t)\right)
\end{equation}
with epoch $T_0$ and duration $D$.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.9\linewidth]{../figures/protopapas.pdf}
        \caption{Simulations of the single non-periodic transit signal $\tau_c(t, T_0, D)$ (\autoref{eq:protopapas_single}) for different values of $c$}
        \label{fig:protopapas}
    \end{centering}
\end{figure}
In our study, we simulate all transit signals with $c=12$. The periodic transit signal $\mathcal{T}$ seen in \autoref{fig:app_principle_dataset} corresponds to $\mathcal{T} = \mathcal{T}_{c=12}(\bm{t}, P=0.7, T_0=0.2, D=0.05)$, all parameters in unit of \textit{days}.

\subsection{Instrumental signals $\mathcal{S}$}
We simulate instrumental signals as a linear model of $M$ explanatory variables arranged in the $(M\times N)$ design matrix $\bm{X}$. Hence,
$$
    \mathcal{S} = \bm{w}\bm{X}
$$
where the vector $\bm{w}$ are linear coefficients of the model. The simulated flux shown in \autoref{fig:app_principle_dataset} contains a linear model where the $M=4$ columns of the design matrix $\bm{X}$ are given by $\bm{X}_{i} = \bm{t}^i$ (i.e. $\bm{X}$ is the Vandermonde matrix order $3$ of time $t$) and $\bm{w} = [1.0\quad0.0005\quad\text{-}0.0002\quad\text{-}0.0005].$

\subsection{Stellar variability $\nu$} 
We simulate correlated noise thanks to Gaussian Processes. In our study, we are particularly interested in stellar variability and its effect on transit detection. For this reason, we use a physically-motivated kernel, describing the covariance of a stochastically-driven damped harmonic oscillator (SHO, \cite{, celerite, celerite2}) taking the form

\begin{equation}
    \begin{gathered}
        k(\tau) = \sigma^2\,\exp\left(-\frac{\omega\,\tau}{2\,Q}\right)
        \left\{\begin{array}{ll}
            1 + \omega\,\tau & \mbox{for } Q = 1/2 \\
            \cosh(f\,\omega\,\tau/2\,Q) + \sinh(f\,\omega\,\tau/2\,Q)/f
                & \mbox{for } Q < 1/2 \\
            \cos(g\,\omega\,\tau/2\,Q) + \sin(g\,\omega\,\tau/2\,Q)/g
                & \mbox{for } Q > 1/2
        \end{array}\right. \\
        \text{where}\quad \tau = |t_i - t_j|\text{,}\quad f = \sqrt{1 - 4\,Q^2} \quad \text{and}\quad g = \sqrt{4\,Q^2 - 1}
    \end{gathered}
\end{equation}
We use this kernel through its implementation in the \texttt{tinygp}\footnote{\href{https://github.com/dfm/tinygp}{https://github.com/dfm/tinygp}} Python package, providing an implententation of the quasi-separable kernel from \citealt{celerite2} powered by \texttt{JAX}\footnote{\href{https://github.com/google/jax}{https://github.com/google/jax}}. The stellar variability signal in \autoref{fig:app_principle_dataset} has been sampled from a Gaussian Process with an SHO kernel of parameters $\omega = \pi/6D$ (i.e. a period equal to 12 times the duration $D$ of the simulated transit), $Q=45$ and $\sigma=\Delta$, the depth of the simulated transit. An extra term $\sigma_f^2=0.001^2$ is added to the covariance matrix, corresponding to the variance of the simulated measurement $f$ and leading to the white noise observed in \autoref{fig:app_principle_dataset}.

\newpage
\section{Combining transits}\label{combining_transits}
